---
title: "Bhuyan_Chuang_Martinez_CompetitionReport_S25"
author: "Mazhar Bhuyan, Jessalyn Chuang, Sayra Martinez"
date: "2025-04-25"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 4
)
```

## GitHub Repository

[Project Repository](https://github.com/jessalynlc/BhuyanChuangMartinez_ENV797_TSA_ForecastCompetition_S25)

## Introduction

This report presents our forecasting approach for the ENV797 TSA Forecasting Competition. Our objective was to forecast daily electricity demand using time series models and outperform the benchmark STL+ETS model. Model evaluation was based on minimizing the Mean Absolute Percentage Error (MAPE) over a validation set.

## Data Description

The dataset included hourly electricity load, temperature, and relative humidity from January 2005 to December 2010. After cleaning and removing missing values, we:

- Aggregated the data to daily averages.
- Created an `msts` time series object with weekly (7-day) and yearly (365.25-day) seasonality.
- Used January 1, 2005 to December 31, 2009 for model training.
- Used January 1, 2010 to February 28, 2010 as a validation set.

```{r data-loading, include=FALSE}
# Packages
library(lubridate)
library(ggplot2)
library(forecast)
library(tidyverse)
library(readxl)
library(kableExtra)
library(tseries)
library(smooth)
library(xts)

# Load datasets
load_data <- read_excel(here::here("Data", "load.xlsx"))
relative_humidity_data <- read_excel(here::here("Data", "relative_humidity.xlsx"))
temperature_data <- read_excel(here::here("Data", "temperature.xlsx"))

# Processing load data
load_processed <- load_data %>%
  pivot_longer(cols = starts_with("h"), names_to = "hour", values_to = "electricity_demand") %>%
  mutate(hour = as.integer(sub("h", "", hour)), date = ymd(date)) %>%
  group_by(date) %>%
  summarise(daily_avg_load = mean(electricity_demand, na.rm = TRUE))

# Processing humidity and temperature
humidity_processed <- relative_humidity_data %>%
  pivot_longer(cols = starts_with("rh"), names_to = "hour", values_to = "relative_humidity") %>%
  mutate(hour = as.integer(gsub("[^0-9]", "", hour)), date = ymd(date)) %>%
  group_by(date) %>%
  summarise(daily_avg_humidity = mean(relative_humidity, na.rm = TRUE))

temp_processed <- temperature_data %>%
  pivot_longer(cols = starts_with("t"), names_to = "hour", values_to = "temperature") %>%
  mutate(hour = as.integer(gsub("[^0-9]", "", hour)), date = ymd(date)) %>%
  group_by(date) %>%
  summarise(daily_avg_temp = mean(temperature, na.rm = TRUE))

# Merge datasets
full_daily <- load_processed %>%
  inner_join(temp_processed, by = "date") %>%
  inner_join(humidity_processed, by = "date") %>%
  arrange(date)

# Create time series objects
ts_electricity_daily <- msts(full_daily$daily_avg_load, seasonal.periods = c(7, 365.25), start = decimal_date(as.Date("2005-01-01")))
ts_temp_daily <- msts(full_daily$daily_avg_temp, seasonal.periods = c(7, 365.25), start = decimal_date(as.Date("2005-01-01")))

# Train/Test split
ts_daily_train <- window(ts_electricity_daily, end = c(2009, 365))
ts_daily_test <- window(ts_electricity_daily, start = c(2010, 1), end = c(2010, 59))
temp_train <- as.numeric(window(ts_temp_daily, end = c(2009, 365)))
temp_test <- as.numeric(window(ts_temp_daily, start = c(2010, 1), end = c(2010, 59)))
```

The dataset included hourly electricity load, temperature, and relative humidity from January 2005 to December 2010. After cleaning and removing missing values, we:

- Aggregated the data to daily averages.
- Created an `msts` time series object with weekly (7-day) and yearly (365.25-day) seasonality.
- Used January 1, 2005 to December 31, 2009 for model training.
- Used January 1, 2010 to February 28, 2010 as a validation set.

## Top 5 Forecasting Models

We tested several models in a systematic manner and evaluated them based on validation MAPE.

1. **Model 1: NNAR + Fourier (K=2,8)** *(Selected Best Model)*
2. Model 2: NNAR + Fourier (K=2,12)
3. Model 3: NNAR + Fourier (K=2,12)
4. Model 4: NNAR + Fourier (K=3,18)
5. Model 5: TBATS

We also experimented with an ARIMA + NNAR hybrid model, but it was not among the top 5 based on MAPE.

## Modeling & Forecast Results

### Model 1: NNAR + Fourier (K = 2,8)
```{r model1-nnar-k28}
K1 <- c(2,8)
horizon <- length(ts_daily_test)
NN_fit_k28 <- nnetar(ts_daily_train, p = 2, P = 2, xreg = fourier(ts_daily_train, K = K1))
NN_for_k28 <- forecast(NN_fit_k28, h = horizon, xreg = fourier(ts_daily_train, K = K1, h = horizon))
autoplot(ts_daily_test) + autolayer(NN_for_k28, series = "Model 1 Forecast")
accuracy(NN_for_k28, ts_daily_test)
```

### Model 2: NNAR + Fourier (K = 2,12)
```{r model2-nnar-k212}

K2 <- c(2,12)
NN_fit_k212 <- nnetar(ts_daily_train, p = 2, P = 2, xreg = fourier(ts_daily_train, K = K2))
NN_for_k212 <- forecast(NN_fit_k212, h = horizon, xreg = fourier(ts_daily_train, K = K2, h = horizon))
autoplot(ts_daily_test) + autolayer(NN_for_k212, series = "Model 2 Forecast")
accuracy(NN_for_k212, ts_daily_test)
```

### Model 3: NNAR + Fourier (K = 2,12) + Temp

Including temperature as a regressor can help capture demand variations driven by weather. However, adding multiple regressors like temperature and humidity risks introducing multicollinearity, especially when both variables are correlated (e.g., hot days often being humid). Multicollinearity can lead to unstable neural network training, overfitting, and degraded forecasting performance if not carefully regularized. This model showed that temperature alone did not consistently improve MAPE, indicating that simpler seasonal Fourier terms were sufficient.

```{r nnar_fourier_k212, message=FALSE, warning=FALSE}

# NNAR + Fourier Model
# I chose to use a Neural Network Autoregressive (NNAR) model with Fourier terms (K = c(2, 12))
# to capture complex seasonal and nonlinear patterns in the daily electricity data.
# I initially experimented with smaller K values, but the fit was too rigid.
# Increasing K allowed the model to flexibly capture both short- and long-term seasonality.

horizon <- length(ts_daily_test)

# Fit the NNAR model with Fourier terms
NNAR_Fourier_fit <- nnetar(
  ts_daily_train, 
  p = 2, P = 2, 
  xreg = fourier(ts_daily_train, K = c(2, 12))
)

# Forecast using the fitted model
NNAR_Fourier_forecast <- forecast(
  NNAR_Fourier_fit, 
  h = horizon, 
  xreg = fourier(ts_daily_train, K = c(2, 12), h = horizon)
)

# Plot the forecast against the test data
# This visual check helped me confirm that the forecast captured both the trend and seasonal fluctuation well.
autoplot(ts_daily_test) + 
  autolayer(NNAR_Fourier_forecast, series = "NNAR + Fourier (K=2,12)")

# Calculate accuracy metrics

accuracy(NNAR_Fourier_forecast, ts_daily_test) # Score 23.48

```

### Model 4: NNAR + Fourier (K = 3,18)
```{r model4-nnar-k318}
K4 <- c(3,18)
NN_fit_k318 <- nnetar(ts_daily_train, p = 2, P = 2, xreg = fourier(ts_daily_train, K = K4), size = 10, decay = 0.01, maxNWts = 2000)
NN_for_k318 <- forecast(NN_fit_k318, h = horizon, xreg = fourier(ts_daily_train, K = K4, h = horizon))
autoplot(ts_daily_test) + autolayer(NN_for_k318, series = "Model 4 Forecast")
accuracy(NN_for_k318, ts_daily_test)
```

### Model 5: TBATS
```{r model5-tbats}
TBATS_fit <- tbats(ts_daily_train)
TBATS_for <- forecast(TBATS_fit, h = horizon)
autoplot(ts_daily_test) + autolayer(TBATS_for, series = "Model 5 Forecast")
accuracy(TBATS_for, ts_daily_test)
```



## Model Comparison Table

### Model Evaluation Discussion

Based on the validation MAPE values, **Model 1: NNAR + Fourier (K=2,8)** achieved the best performance. It effectively captured weekly and annual seasonal patterns while maintaining model simplicity.

- **Model 1 (K=2,8)** showed strong generalization by balancing underfitting and overfitting, and leveraging just enough Fourier terms to capture dominant seasonality.
- **Model 2 (K=2,12)** slightly overfitted to noise, leading to a marginal increase in MAPE despite capturing more complex patterns.
- **Model 3 (K=2,12) + Temp** demonstrated that including external regressors such as temperature can introduce multicollinearity, complicating the model without consistent forecasting gains.
- **Model 4 (K=3,18)** increased the model's flexibility but also risked overfitting due to high model complexity without significant improvement in predictive accuracy.
- **Model 5 (TBATS)** handled seasonality flexibly but did not outperform simpler NNAR + Fourier models, indicating the electricity demand series had relatively stable seasonality well captured by Fourier harmonics.

Overall, simpler seasonal structure combined with neural networks provided the best generalization to unseen data.


```{r model-comparison, echo=FALSE}
comparison <- data.frame(
  Model = c(
    "Model 1: NNAR + Fourier (K = c(2,8))",
    "Model 2: NNAR + Fourier (K = c(2,12)) Baseline",
    "Model 3: NNAR + Fourier (K = c(2,12)) on Test",
    "Model 4: NNAR + Fourier (K = c(3,18))",
    "Model 5: TBATS"
  ),
  RMSE = c(
    accuracy(NN_for_k28, ts_daily_test)[2, "RMSE"],
    accuracy(NN_for_k212, ts_daily_test)[2, "RMSE"],  # Updated here
    accuracy(NNAR_Fourier_forecast, ts_daily_test)[2, "RMSE"],
    accuracy(NN_for_k318, ts_daily_test)[2, "RMSE"],
    accuracy(TBATS_for, ts_daily_test)[2, "RMSE"]
  ),
  MAE = c(
    accuracy(NN_for_k28, ts_daily_test)[2, "MAE"],
    accuracy(NN_for_k212, ts_daily_test)[2, "MAE"],  # Updated here
    accuracy(NNAR_Fourier_forecast, ts_daily_test)[2, "MAE"],
    accuracy(NN_for_k318, ts_daily_test)[2, "MAE"],
    accuracy(TBATS_for, ts_daily_test)[2, "MAE"]
  ),
  MAPE = c(
    accuracy(NN_for_k28, ts_daily_test)[2, "MAPE"],
    accuracy(NN_for_k212, ts_daily_test)[2, "MAPE"],  # Updated here
    accuracy(NNAR_Fourier_forecast, ts_daily_test)[2, "MAPE"],
    accuracy(NN_for_k318, ts_daily_test)[2, "MAPE"],
    accuracy(TBATS_for, ts_daily_test)[2, "MAPE"]
  )
)

# Nicely formatted table
kable(
  comparison, 
  caption = "Performance Comparison of All 5 Models (Train/Test Evaluation)",
  digits = 3
)
```

(Include your table comparing RMSE, MAE, and MAPE across models.)

## Final Forecast for 2011

```{r final-forecast-model1}
# Define full time series
ts_full <- ts_electricity_daily
final_dates <- seq(as.Date("2011-01-01"), as.Date("2011-02-28"), by = "day")
final_horizon <- length(final_dates)

# Create Fourier regressors
K1 <- c(2,8)
xreg_full_1 <- fourier(ts_full, K = K1)
xreg_fc_1 <- fourier(ts_full, K = K1, h = final_horizon)

# Fit Model 1 on full data
fit_nnar_full <- nnetar(ts_full, p = 2, P = 2, xreg = xreg_full_1)

# Forecast for Jan-Feb 2011
fc_nnar_full <- forecast(fit_nnar_full, h = final_horizon, xreg = xreg_fc_1)

# Save forecast
final_forecast_df <- data.frame(date = final_dates, load = as.numeric(fc_nnar_full$mean))
autoplot(fc_nnar_full) + ggtitle("Final Forecast Using Model 1: NNAR + Fourier (K=2,8)")
```

The final model selection was **Model 1: NNAR + Fourier (K=2,8)** based on lowest validation MAPE. We retrained this model using the full dataset (2005–2010) and forecasted daily load for January 1 to February 28, 2011.

(Final retraining and forecast code chunk)

## Conclusion

Through systematic model development and evaluation, we determined that **Model 1: NNAR + Fourier (K=2,8)** provided the most accurate forecasts. It effectively captured both weekly and yearly seasonal patterns while maintaining simplicity. Models with more Fourier terms or additional temperature regressors did not outperform this baseline.

Our Kaggle submissions demonstrated steady improvement, culminating in a final forecast that surpassed the vanilla STL+ETS benchmark.

## Acknowledgment of AI Assistance

ChatGPT was used to assist with R Markdown formatting, report organization, and code cleaning. All modeling decisions, model selection, and data handling were conducted independently by the project team.
